Interpretability and Fairness in Machine Learning: A Formal Methods Approach

### 2023-02-08 doctoral_consortium/main.tex
updated templates for doctoral consortium

\chapter{Introduction}

@@

\section{Tools}
	We develop following open-source tools for interpretable and fair machine learning.
	\begin{itemize}
		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification and FIF computation
	\end{itemize}
\section{Outline}
We organize the thesis in the following way. We discuss preliminaries in Chapter~\ref{chapter:preliminaries}. In the first part of the thesis, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable classifiers. We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}. In the second part of this thesis, we discuss  fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}. Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.

\begin{comment}

The tremendous success of machine learning in recent years has found its applications in different high-stake and  safety-critical domains, such as medical, law, education, and transportation. In such domains, the interpretability and fairness of machine learning classifiers are of paramount importance. In this work, we focus on improving the interpretability and fairness in machine learning.

There are two-fold objectives in interpretable machine learning: (i) designing classifiers that are interpretable by design and (ii) leveraging already interpretable models to explain the prediction of black-box classifiers. In this work, we propose an efficient learning framework~\cite{ghosh19incremental} for a rule-based classifier\textemdash a popular interpretable model because of expressing the decision function in terms of rules corresponding to input features.  Our framework allows us to learn classification rules expressible in propositional logic, for example, CNF (Conjunctive Normal Form) and DNF (Disjunctive formal form) formulas.  The benefit of learning CNF classifiers is that this is the fundamental block for learning other popular interpretable representations: decision lists and decision sets. Learning rule-based classifiers is an intractable problem because of solving a combinatorial optimization problem. To efficiently learn these classifiers, we propose a MaxSAT-based formulation and scale it by combining incremental learning. The result is an efficient learning framework that can classify datasets even with 1 million samples while also balancing between high accuracy and small rule size. As an application of incremental learning, we propose to learn more relaxed logical rules~\cite{ghosh20classification} which demonstrate higher succinctness in their representation while achieving better accuracy. Furthermore, as an application of MaxSAT, we solve group testing problems based on compressed sensing~\cite{ciampiconi20maxsat}. Our MaxSAT-based formulation for group testing has impressive scalability than state-of-the-art approaches.

We study the application of interpretable machine learning models in explaining the prediction of black-box models such as neural networks. To this end, we propose a PAC-based (Probably Approximately Correct) local explanations of feed-forward and recurrent neural networks using rule-based classifiers and their variants~\cite{neider2020probably,ghosh2020formal}. To explain feed-forward neural networks, we learn flexible classification rules using SyGuS~\cite{neider2020probably}. Such rules have the theoretical property that they are PAC approximate to the neural network within the locality of a query of interest. Similarly, to explain recurrent neural networks, we propose learning explanations in LTL formulas (Linear Temporal Logic)~\cite{ghosh2020formal}.

Machine learning models may become biased in their decisions towards certain demographic populations. One important question in fairness in machine learning is to verify whether the model is unfair or not with respect to multiple definitions of fairness studied in the literature. To this end, we propose a scalable fairness verification framework based on formal methods, particularly stochastic satisfiability (SSAT)~\cite{ghosh2020justicia} and stochastic subset-sum problem (S3P)~\cite{ghosh2021algorithmic}. The input to the fairness verifier is a machine learning model and the distribution of input features and the output is an assessment of different fairness metrics that the model obtains given the distribution. Our fairness verifier has the following advantages over existing verifiers. We can efficiently verify the fairness of compound sensitive groups\textemdash such as, white-male and black-female\textemdash by combining multiple sensitive features, unlike existing works that only consider a single binary sensitive feature. We also consider feature-correlation represented as a Bayesian network whereas existing methods consider limited correlation among features. As a result, the proposed verifier has higher accuracy than existing verifiers. Finally, our verifier is more scalable than existing verifiers because of the efficient formulation based on SSAT and S3P.

As future work, we aim to combine the interpretability and fairness of machine learning in a single framework. In particular, we explain the unfairness of models by combining explanation methods in machine learning. Our hypothesis is that such a combination would be effective in identifying the bias of a model and also mitigate the bias through imperative actions.

\end{comment}

>>>>

%%%% ijcai23.tex

\typeout{IJCAI--23 Instructions for Authors}

% These are the instructions for authors for IJCAI-23.

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in

% The file ijcai23.sty is a copy from ijcai22.sty
% The file ijcai22.sty is NOT the same as previous years'
\usepackage{ijcai23}

% Use the postscript times font!
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[switch]{lineno}

% Comment out this line in the camera-ready submission
\linenumbers

\urlstyle{same}

% the following package is optional:
%\usepackage{latexsym}

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}

% Following comment is from ijcai97-submit.tex:
% The preparation of these files was supported by Schlumberger Palo Alto
% Research, AT\&T Bell Laboratories, and Morgan Kaufmann Publishers.
% Shirley Jowell, of Morgan Kaufmann Publishers, and Peter F.
% Patel-Schneider, of AT\&T Bell Laboratories collaborated on their
% preparation.

% These instructions can be modified and used in other conferences as long
% as credit to the authors and supporting agencies is retained, this notice
% is not changed, and further modification or reuse is not restricted.
% Neither Shirley Jowell nor Peter F. Patel-Schneider can be listed as
% contacts for providing assistance without their prior permission.

% To use for other conferences, change references to files and the
% conference appropriate and use other authors, contacts, publishers, and
% organizations.
% Also change the deadline and address for returning papers and the length and
% page charge instructions.
% Put where the files are available in the appropriate places.

% PDF Info Is REQUIRED.
% Please **do not** include Title and Author information
\pdfinfo{
/TemplateVersion (IJCAI.2023.0)
}

\title{Interpretability and Fairness in Machine Learning: A Formal Methods Approach}

% Single author syntax
\author{
    Bishwamittra Ghosh
    \affiliations
    National University of Singapore
    \emails
   bghosh@u.nus.edu
}

% Multiple author syntax (remove the single-author syntax above and the \iffalse ... \fi here)
\iffalse
\author{
First Author$^1$
\and
Second Author$^2$\and
Third Author$^{2,3}$\And
Fourth Author$^4$
\affiliations
$^1$First Affiliation\\
$^2$Second Affiliation\\
$^3$Third Affiliation\\
$^4$Fourth Affiliation
\emails
\{first, second\}@example.com,
third@other.example.com,
fourth@example.com
}
\fi

% Macros
\newcommand{\imli}{\ensuremath{\mathsf{IMLI}}}
\newcommand{\crr}{\ensuremath{\mathsf{CRR}}}
\newcommand{\fairXplainer}{\ensuremath{\mathsf{FairXplainer}}}
\newcommand{\justicia}{\ensuremath{\mathsf{Justicia}}}
\newcommand{\fvgm}{\ensuremath{\mathsf{FVGM}}}

\begin{document}

\maketitle

@@

%% The file named.bst is a bibliography style file for BibTeX 0.99c
\bibliographystyle{named}
\bibliography{ref}

\end{document}
